Simple repression draft

### 2020-10-17 doc/resubmission/appendix_01_nonbursty.tex
minor fixes to typos

The chemical master equation describes the continuous time evolution of a
continuous or discrete probability distribution function. In our specific case
we want to describe the time evolution of the discrete mRNA distribution. What
this means is that we want to compute the probability of having $m$ mRNA
molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
interval such that only one of the possible reactions take place during that
time interval. For the example that we will work out here in detail we chose the
two-state stochastic simple repression model schematized in
Figure~\ref{fig:two_state}(A). To derive the master equation we will focus more
on the representation shown in Figure~\ref{fig:two_state}(B), where the
transitions between different mRNA counts and promoter states is more explicitly
depicted. Given that the DNA promoter can exist in one of two states --
transcriptionally active state, and with repressor bound -- we will keep track
not only of the mRNA count, but on which state the promoter is. For this we will
keep track of two probability distributions: The probability of having $m$ mRNAs
at time $t$ when the promoter is in the transcriptionally active state $A$,
$p_A(m, t)$, and the equivalent probability but when the promoter is in the
repressor bound state $R$, $p_R(m, t)$.

>>>>

The chemical master equation describes the continuous time evolution of a
continuous or discrete probability distribution function. In our specific case
we want to describe the time evolution of the discrete mRNA distribution. What
this means is that we want to compute the probability of having $m$ mRNA
molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
interval such that only one of the possible reactions take place during that
time interval. For the example that we will work out here in detail we chose the
two-state stochastic simple repression model schematized in
Figure~\ref{fig:two_state}(A). To derive the master equation we will focus more
on the representation shown in Figure~\ref{fig:two_state}(B), where the
transitions between different mRNA counts and promoter states is more explicitly
depicted. Given that the DNA promoter can exist in one of two states--
transcriptionally active state, and with repressor bound-- we will keep track
not only of the mRNA count, but on which state the promoter is. For this we will
keep track of two probability distributions: The probability of having $m$ mRNAs
at time $t$ when the promoter is in the transcriptionally active state $A$,
$p_A(m, t)$, and the equivalent probability but when the promoter is in the
repressor bound state $R$, $p_R(m, t)$.

### 2020-10-17 doc/resubmission/appendix_02_bursty.tex
minor fixes to typos

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'={m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization constraint of the geometric distribution, adding the
probability of all possible burst sizes -- including size zero since we
re-indexed the sum -- allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

>>>>

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'={m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization constraint of the geometric distribution, adding the
probability of all possible burst sizes --including size zero since we
re-indexed the sum-- allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

### 2020-10-17 doc/resubmission/appendix_03_bayesian_inference.tex
minor fixes to typos

One could argue that the whole goal of formulating theoretical models about
nature is to sharpen our understanding from qualitative statements to precise
quantitative assertions about the relevant features of the natural phenomena in
question \cite{Gunawardena2014}. It is in these models that we intend to distill
the essential parts of the object of study. Writing down such models leads to a
propagation of mathematical variables that parametrize our models. By assigning
numerical values to these parameters we can compute concrete predictions that
can be contrasted with experimental data. For these predictions to match the
data the parameter values have to carefully be chosen from the whole parameter
space. But how do we go about assessing the effectiveness of different regions
of parameter space to speak to the ability of our model to reproduce the
experimental observations? The language of probability, and more specifically of
Bayesian statistics is -- we think -- the natural language to tackle this
question.

@@

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given our observation of some
particular dataset. In other words, through the application of Bayes' formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is captured in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes' formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes' theorem, we limit 
ourselves to the terms that depend on our quantity of interest -- that is the 
parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
serves as a normalization constant.

@@

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes' theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is not uncommon practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes' theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomenon in the form of a mathematical model -- in the context of Bayesian
inference this means choosing a form for the likehihoods, and even making
this choice presupposes prior understanding or assumptions as to the
relevant features in the system under study -- but that we have absolutely
no idea what the parameter values could or could not be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down weakly-informative prior distributions
for our parameters.

>>>>

One could argue that the whole goal of formulating theoretical models about
nature is to sharpen our understanding from qualitative statements to precise
quantitative assertions about the relevant features of the natural phenomena in
question \cite{Gunawardena2014}. It is in these models that we intend to distill
the essential parts of the object of study. Writing down such models leads to a
propagation of mathematical variables that parametrize our models. By assigning
numerical values to these parameters we can compute concrete predictions that
can be contrasted with experimental data. For these predictions to match the
data the parameter values have to carefully be chosen from the whole parameter
space. But how do we go about assessing the effectiveness of different regions
of parameter space to speak to the ability of our model to reproduce the
experimental observations? The language of probability, and more specifically of
Bayesian statistics is --we think-- the natural language to tackle this
question.

@@

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given our observation of some
particular dataset. In other words, through the application of Bayes' formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is captured in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes' formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes' theorem, we limit 
ourselves to the terms that depend on our quantity of interest --that is the 
parameter values themselves $\vec{\theta}$-- as the denominator $P(D)$ only
serves as a normalization constant.

@@

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes' theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is not uncommon practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes' theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomenon in the form of a mathematical model --in the context of Bayesian
inference this means choosing a form for the likehihoods, and even making
this choice presupposes prior understanding or assumptions as to the
relevant features in the system under study-- but that we have absolutely
no idea what the parameter values could or could not be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down weakly-informative prior distributions
for our parameters.

### 2020-10-17 doc/resubmission/section_02_means.tex
minor fixes to typos

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no
transcription for many of these states. This discretization of a potentially
continuous number of promoter states (due to effects such as supercoiling of
DNA~\cite{Chong2014, Sevier2016} or DNA looping \cite{Boedicker2013a}) is
analogous to how the Monod-Wyman-Changeux model of allostery coarse-grains
continuous molecule conformations into a finite number of
states~\cite{Martins2011}. The models are probabilistic with each state assigned
some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key aspects: what states
to consider and how to compute the probabilities of those states.

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel out.

Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
mathematicized cartoons for simple repression using both thermodynamic and
kinetic models, respectively, that have appeared in previous literature. For
each cartoon, we calculate the fold-change in mean gene expression as predicted
by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
fold-change can be written as a Fermi function of the form
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~~\ref{fig1:means_cartoons}(D), left panel). This implies all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.

>>>>

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription --including no
transcription for many of these states. This discretization of a potentially
continuous number of promoter states (due to effects such as supercoiling of
DNA~\cite{Chong2014, Sevier2016} or DNA looping \cite{Boedicker2013a}) is
analogous to how the Monod-Wyman-Changeux model of allostery coarse-grains
continuous molecule conformations into a finite number of
states~\cite{Martins2011}. The models are probabilistic with each state assigned
some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key aspects: what states
to consider and how to compute the probabilities of those states.

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel out. Figure~\ref{fig1:means_cartoons}(B) and
(C) depicts a smorgasbord of mathematicized cartoons for simple repression using
both thermodynamic and kinetic models, respectively, that have appeared in
previous literature. For each cartoon, we calculate the fold-change in mean gene
expression as predicted by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
fold-change can be written as a Fermi function of the form
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~~\ref{fig1:means_cartoons}(D), left panel). This implies all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.
